---
title: "Preprocessing"
output: html_document
---

# Preprocessing

1.  Libraries and data upload
2.  Raw data analysis
3.  Data preparation and summaries
4.  Log transformation
5.  Stationarity
6.  ACF and PACF
7.  First Difference (7.1 Stationarity; 7.2 ACF and PACF)
8.  Training and Test Sets

## 1. Libraries and data upload

```{r}
# 1. Libraries and data upload 
library(ggplot2)
library(urca)
library(tseries)
data <- read.csv("data/UMCSENT.csv")

```

## 2. Raw data analysis

```{r}
# 2. Raw data analysis
str(data) # observation_date chr; UMCSENT num
data$observation_date <- as.Date(data$observation_date, format = "%Y-%m-%d") # observation_date conversion to Date
str(data)

sum(is.na(data$UMCSENT)) # 210 missing values 
mean(is.na(data$UMCSENT)) # 24% of missing values

nrow(data) # 875
range(data$observation_date) # "1952-11-01" "2025-09-01"

ggplot(data, aes(x = observation_date, y = UMCSENT)) +
  geom_line() +
  labs(title = 'Consumer Sentiment Index')

```

## 3. Data preparation and summaries

```{r}

# 3. Data preparation and summaries

# data from 1978
data_filtered <- subset(data, observation_date >= "1978-01-01")
nrow(data_filtered) # 573
range(data_filtered$observation_date) # "1978-01-01" "2025-09-01"

ggplot(data_filtered, aes(x = observation_date, y = UMCSENT)) +
  geom_line() +
  labs(title = 'Consumer Sentiment Index')

# statistics
summary(data_filtered$UMCSENT) 

# check outliers
hist(data_filtered$UMCSENT, breaks = 20) # not normally distributed
boxplot(data_filtered$UMCSENT) # values are distributed in the 70-90 range with 88.2 as the median


```

## 4. Log transformation

```{r}

# Log transformation for positive time series
data_filtered$log_UMCSENT <- log(data_filtered$UMCSENT)

ggplot(data_filtered, aes(x = observation_date, y = log_UMCSENT)) +
  geom_line() +
  labs(title = 'Log (Consumer Sentiment Index)')

```

## 5. Stationarity

```{r}
# Stationarity test
# ADF -> stationary
summary(ur.df(data_filtered$log_UMCSENT, type = "drift"))
# KPSS -> non-stationary
kpss.test(data_filtered$log_UMCSENT)

```

### 1. Augmented Dickey-Fuller (ADF) Test

This test checks for the presence of a unit root (non-stationarity).

-   $H_0$ (Null Hypothesis): The series is **not stationary** (it has a unit root).

-   $H_1$ (Alternative Hypothesis): The series **is stationary**.

-   **Result:** The test statistic value ($t$) is **-3.5248**. This is more negative than the 1% critical value (-3.43).

-   **Conclusion:** **Reject** $H_0$. According to this test, the series is considered **stationary** (with a 99% confidence level).

### 2. KPSS Test

This test works with the opposite hypotheses compared to the ADF test.

-   $H_0$ (Null Hypothesis): The series **is stationary** (level stationarity).

-   $H_1$ (Alternative Hypothesis): The series is **not stationary**.

-   **Result:** The *p-value* is **0.01** (indicating it is smaller than the printed value).

-   **Conclusion:** Since the p-value is lower than the standard threshold of 0.05 (5%), **reject** $H_0$. According to this test, the series is **not stationary**.

## 6. Autocorrelation

```{r}
# ACF and PACF
acf(data_filtered$log_UMCSENT)
pacf(data_filtered$log_UMCSENT)

```

## 7. First difference

```{r}
# First difference
diff_log <- diff(data_filtered$log_UMCSENT)
data_filtered$diff_log_UMCSENT <- c(NA, diff_log)

ggplot(data_filtered, aes(x = observation_date, y = diff_log_UMCSENT)) +
  geom_line() +
  labs(title = 'First difference log(Consumer Sentiment Index)')

```

### 7.1 Stationarity

```{r}

# Stationarity test on differenced data
# ADF -> stationary
summary(ur.df(diff_log, type = "drift"))
# KPSS -> non-stationary
kpss.test(diff_log)
```

### 1. Augmented Dickey-Fuller (ADF) Test

-   $H_0$ (Null Hypothesis): The series is **not stationary** (has a unit root).

-   $H_1$ (Alternative Hypothesis): The series **is stationary**.

-   **Result:** The test statistic is **-19.6373**. This is extremely negative and far below the 1% critical value (-3.43).

-   **Conclusion:** **Strongly reject** $H_0$. The series is definitely **stationary**.

### 2. KPSS Test

-   $H_0$ (Null Hypothesis): The series **is stationary**.

-   $H_1$ (Alternative Hypothesis): The series is **not stationary**.

-   **Result:** The *p-value* is **0.1** (with the warning "p-value greater than printed p-value," implying the actual probability is even higher, likely \> 0.10).

-   **Conclusion:** Since the p-value is greater than 0.05, **fail to reject** $H_0$. The evidence supports the conclusion that the series is **stationary**.

### 7.2 ACF and PACF

```{r}

# ACF and PACF of differenced data
par(mfrow = c(1, 2)) 

# Autocorrelation Function (per stimare q - parte MA)
acf(diff_log, main = "ACF - First Difference", lag.max = 36)

# Partial Autocorrelation Function (per stimare p - parte AR)
pacf(diff_log, main = "PACF - First Difference", lag.max = 36)

par(mfrow = c(1, 1))

# ACF values
acf_values <- acf(diff_log, lag.max = 36, plot = FALSE)

acf_table <- data.frame(
  Lag = acf_values$lag[,,1], 
  ACF = acf_values$acf[,,1]
)
print(acf_table)

# PACF values
pacf_values <- pacf(diff_log, lag.max = 36, plot = FALSE)

pacf_table <- data.frame(
  Lag = pacf_values$lag[,,1], 
  PACF = pacf_values$acf[,,1]
)
print(pacf_table)

```

### 1. General Interpretation

The most striking feature of both tables is that **Lag 1 is statistically insignificant** (approx 0.03), while **Lag 2 is significant** (approx -0.17).

-   **Absence of Short-Term Memory:** The fact that Lag 1 is near zero in both ACF and PACF means the series does not depend on the immediately preceding month.

-   **Lag 2 Dependency:** The significant negative spike at Lag 2 in both functions suggests a specific 2-period relationship.

Because the ACF and PACF look somewhat similar (both have a spike at Lag 2 and are otherwise quiet), it indicates a process that can be modeled effectively by either pure AR or pure MA terms, provided they reach the second lag.

### 2. Best Possible Models (Candidates)

Based on the standard identification rules (Box-Jenkins methodology), here are the top candidates:

#### **Candidate A: ARIMA(0, 1, 2)**

-   **Why?** This relies on the **ACF** interpretation. The ACF shows significant activity up to Lag 2 and then largely cuts off (ignoring the minor Lag 5 noise). This suggests a Moving Average (MA) process of order $q=2$.

-   **Note:** Since Lag 1 is insignificant, the model will likely estimate a very small coefficient for the first MA term ($ma1 \approx 0$) and a significant one for the second ($ma2 \approx -0.17$).

#### **Candidate B: ARIMA(2, 1, 0)**

-   **Why?** This relies on the **PACF** interpretation. The PACF shows significant activity at Lag 2 and then mostly cuts off (again, barring the Lag 5 spike). This suggests an AutoRegressive (AR) process of order $p=2$.

-   **Note:** Similar to the model above, this would likely result in a model where the $ar1$ coefficient is near zero, but $ar2$ is significant.

## 8. Training and Test sets

```{r}

# split dataset into training/test
nrows <- nrow(data_filtered)
index_division <- floor(0.80 * nrows)

train <- subset(data_filtered, 1:nrows <= index_division)
test <- subset(data_filtered, 1:nrows > index_division)

cat("Training:", nrow(train), "rows\n")
cat("Test:", nrow(test), "rows\n")
```

```{r}

library(forecast)

# --- 3 MODELLI ---

# 1. ARIMA(2,1,0) - "La scelta del PACF"
# Include 2 termini autoregressivi per catturare quel picco al lag 2.
model_ar2 <- Arima(train$log_UMCSENT, order = c(2, 1, 0))

# 2. ARIMA(0,1,2) - "La scelta dell'ACF"
# Se leggiamo l'ACF come "significativo ai primi lag e poi zero", proviamo un MA(2).
# Questo modello assume che gli shock (errori) passati influenzino il presente.
model_ma2 <- Arima(train$log_UMCSENT, order = c(0, 1, 2))

# 3. ARIMA(1,1,1) - "Il compromesso"
# Spesso batte modelli più complessi perché è più semplice.
model_mixed <- Arima(train$log_UMCSENT, order = c(1, 1, 1))

# --- CONFRONTO ---

comparison <- data.frame(
  Model = c("ARIMA(2,1,0)", "ARIMA(0,1,2)", "ARIMA(1,1,1)"),
  AIC = c(model_ar2$aic, model_ma2$aic, model_mixed$aic),
  BIC = c(model_ar2$bic, model_ma2$bic, model_mixed$bic),
  # Aggiungiamo l'errore quadratico medio (RMSE) sul training set
  RMSE = c(accuracy(model_ar2)[2], accuracy(model_ma2)[2], accuracy(model_mixed)[2])
)

# Ordiniamo dal migliore (AIC più basso)
comparison <- comparison[order(comparison$AIC),]
print(comparison)

# --- VALIDAZIONE DEL MIGLIOR MODELLO ---
# Prendiamo il primo della lista
best_model_name <- comparison$Model[1]
cat("The best model is:", best_model_name, "\n")

if(best_model_name == "ARIMA(2,1,0)") final_model <- model_ar2
if(best_model_name == "ARIMA(0,1,2)") final_model <- model_ma2
if(best_model_name == "ARIMA(1,1,1)") final_model <- model_mixed

# Controllo finale: I residui sono White Noise?
# Guarda il p-value del Ljung-Box test nel grafico: deve essere > 0.05
checkresiduals(final_model)
```

```{r}
# # CORRELATION TEST
# #Barletts' bands
# bartlett.test(data_filtered$log_UMCSENT ~ as.factor(cut(seq_along(data_filtered$log_UMCSENT), 10)))
# #Box-Pierce
# Box.test(data_filtered$log_UMCSENT, lag = 10, type = "Box-Pierce") # high autocorrelation
# Box.test(diff_log, lag = 10, type = "Box-Pierce")
# #Ljung-Box
# Box.test(data_filtered$log_UMCSENT, lag = 10, type = "Ljung-Box") # high autocorrelation

```
