---
title: "Preprocessing"
output: html_document
---

# Preprocessing

1.  Libraries and data upload
2.  Raw data analysis
3.  Data preparation and summaries
4.  Log transformation
5.  Stationarity
6.  ACF and PACF
7.  First Difference (7.1 Stationarity; 7.2 ACF and PACF)
8.  Training and Test Sets

## 1. Libraries and data upload

```{r}
# 1. Libraries and data upload 
library(ggplot2)
library(urca)
library(tseries)
data <- read.csv("data/UMCSENT.csv")

```

## 2. Raw data analysis

```{r}
# 2. Raw data analysis
str(data) # observation_date chr; UMCSENT num
data$observation_date <- as.Date(data$observation_date, format = "%Y-%m-%d") # observation_date conversion to Date
str(data)

sum(is.na(data$UMCSENT)) # 210 missing values 
mean(is.na(data$UMCSENT)) # 24% of missing values

nrow(data) # 875
range(data$observation_date) # "1952-11-01" "2025-09-01"

ggplot(data, aes(x = observation_date, y = UMCSENT)) +
  geom_line() +
  labs(title = 'Consumer Sentiment Index')

```

## 3. Data preparation and summaries

```{r}

# 3. Data preparation and summaries

# Data from 1978
data_filtered <- subset(data, observation_date >= "1978-01-01")
nrow(data_filtered) # 573
range(data_filtered$observation_date) # "1978-01-01" "2025-09-01"

ggplot(data_filtered, aes(x = observation_date, y = UMCSENT)) +
  geom_line() +
  labs(title = 'Consumer Sentiment Index (1978 - 2025)',
       x = "Year")

# Statistics
summary(data_filtered$UMCSENT) 

# Check outliers

# Histogram
ggplot(data_filtered, aes(x = UMCSENT)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  labs(title = "Consumer Sentiment Index Distribution",
       x = "UMCSENT",
       y = "Frequency") +
  theme_minimal() # not normally distributed

# Boxplot
ggplot(data_filtered, aes(y = UMCSENT)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.colour = "red") +
  labs(title = "Boxplot - Consumer Sentiment Index",
       y = "UMCSENT") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())


```

## 4. Log transformation

```{r}

# Log transformation for positive time series
data_filtered$log_UMCSENT <- log(data_filtered$UMCSENT)

ggplot(data_filtered, aes(x = observation_date, y = log_UMCSENT)) +
  geom_line() +
  labs(title = 'Log (Consumer Sentiment Index)',
       x = "Year")

```

## 5. Stationarity

```{r}
# Stationarity test
# ADF -> stationary
summary(ur.df(data_filtered$log_UMCSENT, type = "drift"))
# KPSS -> non-stationary
kpss.test(data_filtered$log_UMCSENT)

```

### 1. Augmented Dickey-Fuller (ADF) Test

This test checks for the presence of a unit root (non-stationarity).

-   $H_0$ (Null Hypothesis): The series is **not stationary** (it has a unit root).

-   $H_1$ (Alternative Hypothesis): The series **is stationary**.

-   The test statistic value ($t$) is **-3.5248** \< 1% critical value (-3.43).

-   Reject H0 (non-stationarity) -\> the series is **stationary**.

### 2. KPSS Test

-   $H_0$ (Null Hypothesis): The series **is stationary** (level stationarity).

-   $H_1$ (Alternative Hypothesis): The series is **not stationary**.

-   The *p-value* is **0.01** (indicating it is smaller than the printed value) \< 0.05 (5%).

-   Reject $H_0$. According to this test, the series is **not stationary**.

## 6. ACF and PACF

```{r}
# ACF and PACF

# ACF
acf_obj <- acf(data_filtered$log_UMCSENT, plot = FALSE)
acf_df <- data.frame(lag = acf_obj$lag[-1], acf = acf_obj$acf[-1]) 

ci <- 1.96 / sqrt(length(data_filtered$log_UMCSENT))

ggplot(acf_df, aes(x = lag, y = acf)) +
  geom_hline(yintercept = 0, color = "black") + # Asse orizzontale
  geom_hline(yintercept = c(-ci, ci), linetype = "dashed", color = "darkgrey") + 
  geom_segment(aes(xend = lag, yend = 0), color = "black", size = 1) +
  geom_point(size = 2, color = "black") +
  scale_x_continuous(breaks = seq(1, 25, 2), expand = c(0,0.1)) +
  labs(title = "Autocorrelation Function (ACF)", y = "ACF") +
  theme_minimal()

# PACF
pacf_obj <- pacf(data_filtered$log_UMCSENT, plot = FALSE)
pacf_df <- data.frame(lag = pacf_obj$lag, pacf = pacf_obj$acf)

ggplot(pacf_df, aes(x = lag, y = pacf)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_hline(yintercept = c(-ci, ci), linetype = "dashed", color = "darkgrey") +
  geom_segment(aes(xend = lag, yend = 0), color = "black", size = 1) +
  geom_point(size = 2, color = "black") +
  scale_x_continuous(breaks = seq(1, 25, 2), expand = c(0,0.1)) +
  labs(title = "Partial Autocorrelation Function (PACF)", y = "PACF") +
  theme_minimal()


```

## 7. First difference

```{r}
# First difference
diff_log <- diff(data_filtered$log_UMCSENT)
data_filtered$diff_log_UMCSENT <- c(NA, diff_log)

ggplot(data_filtered, aes(x = observation_date, y = diff_log_UMCSENT)) +
  geom_line() +
  labs(title = 'First Difference of Consumer Sentiment Index (in logs)',
       x = "Year",
       y = "Differenced UMCSENT (in logs)")

```

### 7.1 Stationarity Tests on Differenced Data

```{r}

# Stationarity test on differenced data
# ADF -> stationary
summary(ur.df(diff_log, type = "drift"))
# KPSS -> non-stationary
kpss.test(diff_log)
```

### 1. Augmented Dickey-Fuller (ADF) Test

-   The test statistic is **-19.6373** \<1% critical value (-3.43).

-   **reject** $H_0$ -\> series is **stationary**.

### 2. KPSS Test

-   The *p-value* is **0.1** (with the warning "p-value greater than printed p-value," implying the actual probability is even higher, likely \> 0.10) \> 0.05

-   **fail to reject** $H_0$ -\> series is **stationary**.

### 7.2 ACF and PACF - Differenced Data

```{r}

# ACF and PACF of differenced data
par(mfrow = c(1, 2)) 

# Autocorrelation Function - MA
acf(diff_log, main = "ACF - First Difference", lag.max = 36)

# Partial Autocorrelation Function - AR
pacf(diff_log, main = "PACF - First Difference", lag.max = 36)

par(mfrow = c(1, 1))

# ACF values
acf_values <- acf(diff_log, lag.max = 36, plot = FALSE)

acf_table <- data.frame(
  Lag = acf_values$lag[,,1], 
  ACF = acf_values$acf[,,1]
)
print(acf_table)

# PACF values
pacf_values <- pacf(diff_log, lag.max = 36, plot = FALSE)

pacf_table <- data.frame(
  Lag = pacf_values$lag[,,1], 
  PACF = pacf_values$acf[,,1]
)
print(pacf_table)

# ACF

library(ggplot2)
library(gridExtra) 

diff_data_clean <- na.omit(diff_log)

n_obs <- length(diff_data_clean)
ci_threshold <- 1.96 / sqrt(n_obs)

acf_obj <- acf(diff_data_clean, lag.max = 36, plot = FALSE)

acf_df <- data.frame(
  lag = acf_obj$lag[-1], 
  acf_value = acf_obj$acf[-1]
)

plot_acf <- ggplot(acf_df, aes(x = lag, y = acf_value)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_hline(yintercept = c(-ci_threshold, ci_threshold), linetype = "dashed", color = "blue") +
  geom_segment(aes(xend = lag, yend = 0), color = "black", size = 0.7) +
  geom_point(size = 2, color = "black") + 
  scale_x_continuous(breaks = seq(0, 36, 4)) +
  labs(title = "ACF - First Difference", y = "ACF", x = "Lag") +
  theme_minimal()

# PACF
pacf_obj <- pacf(diff_data_clean, lag.max = 36, plot = FALSE)

pacf_df <- data.frame(
  lag = pacf_obj$lag, 
  pacf_value = pacf_obj$acf
)

plot_pacf <- ggplot(pacf_df, aes(x = lag, y = pacf_value)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_hline(yintercept = c(-ci_threshold, ci_threshold), linetype = "dashed", color = "blue") +
  geom_segment(aes(xend = lag, yend = 0), color = "black", size = 0.7) +
  geom_point(size = 2, color = "black") + 
  scale_x_continuous(breaks = seq(0, 36, 4)) +
  labs(title = "PACF - First Difference", y = "PACF", x = "Lag") +
  theme_minimal()

grid.arrange(plot_acf, plot_pacf, ncol = 2)


```

## 8. Training and Test sets

```{r}

# split dataset into training/test
nrows <- nrow(data_filtered)
index_division <- floor(0.80 * nrows)

train <- subset(data_filtered, 1:nrows <= index_division)
test <- subset(data_filtered, 1:nrows > index_division)

cat("Training:", nrow(train), "rows\n")
cat("Test:", nrow(test), "rows\n")
```

## 9. ACF and PACF - Train Set

```{r}

par(mfrow=c(1,2))

# ACF plot - train set
acf(train$diff_log_UMCSENT, na.action = na.pass, main="ACF - Training Data")

# PACF plot - train set
pacf(train$diff_log_UMCSENT, na.action = na.pass, main="PACF - Training Data")

par(mfrow=c(1,1))



```

## 10. Model Selection (AIC/BIC Comparison)

```{r}

library(tidyr) 
library(forecast)

results <- data.frame(Order = integer(),
                      Type = character(),
                      AIC = numeric(),
                      BIC = numeric())

# Models AR and MA of order 1 to 12

for (i in 1:12) {
  
  # Model AR(i): ARIMA(i, 1, 0)
  try({
    fit_ar <- Arima(train$log_UMCSENT, order = c(i, 1, 0))
    results <- rbind(results, data.frame(Order = i, Type = "AR Model", AIC = fit_ar$aic, BIC = fit_ar$bic))
  }, silent = TRUE)
  
  # Model MA(i): ARIMA(0, 1, i)
  try({
    fit_ma <- Arima(train$log_UMCSENT, order = c(0, 1, i))
    results <- rbind(results, data.frame(Order = i, Type = "MA Model", AIC = fit_ma$aic, BIC = fit_ma$bic))
  }, silent = TRUE)
}

results_long <- pivot_longer(results, cols = c("AIC", "BIC"), names_to = "Metric", values_to = "Value")

# Plot
ggplot(results_long, aes(x = Order, y = Value, color = Metric, group = Metric)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ Type, scales = "free_y") + 
  scale_x_continuous(breaks = 1:12) +
  labs(title = "AIC & BIC Comparison by Model Order",
       subtitle = "AR(p) models and MA(q) models",
       x = "Order (p or q)",
       y = "Information Criterion Value",
       color = "Metric") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold")
  )

```

```{r}

# Mixed Models Comparison

library(ggplot2)
library(tidyr)
library(forecast)

manual_models <- list(
  c(1, 1),
  c(1, 2),
  c(2, 1),
  c(2, 2)
)

results_mixed <- data.frame(Model = character(),
                            AIC = numeric(),
                            BIC = numeric(),
                            stringsAsFactors = FALSE)

for (ord in manual_models) {
  p <- ord[1]
  q <- ord[2]
  model_name <- paste0("ARIMA(", p, ",1,", q, ")")
  
  try({
    fit <- Arima(train$log_UMCSENT, order = c(p, 1, q))
    results_mixed <- rbind(results_mixed, data.frame(Model = model_name, AIC = fit$aic, BIC = fit$bic))
  }, silent = TRUE)
}

level_order <- c("ARIMA(1,1,1)", "ARIMA(1,1,2)", "ARIMA(2,1,1)", "ARIMA(2,1,2)")
results_mixed$Model <- factor(results_mixed$Model, levels = level_order)

results_long <- pivot_longer(results_mixed, cols = c("AIC", "BIC"), names_to = "Metric", values_to = "Value")

ggplot(results_long, aes(x = Model, y = Value, color = Metric, group = Metric)) +
  geom_line(size = 0.8) +
  geom_point(shape = 21, size = 3, fill = "white", stroke = 1.2) + 
  scale_color_manual(values = c("AIC" = "salmon", "BIC" = "turquoise")) +
  labs(title = "Information criteria for ARIMA(p,1,q) models",
       x = "",
       y = "") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    legend.position = c(0.9, 0.9),
    legend.box.background = element_rect(color = "black", size = 0.5), 
    legend.title = element_blank(), 
    panel.grid.major = element_line(color = "gray85"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 11, color = "black"),
    axis.text.y = element_text(size = 11, color = "black")
  )

```

### Model comparison

-   ARIMA(0,1,3)

-   ARIMA(1,1,2)

-   ARIMA(2,1,1)

-   Random Walk

```{r}

library(forecast)
library(ggplot2)

full_series <- data_filtered$log_UMCSENT
nrows <- nrow(data_filtered)
index_division <- floor(0.80 * nrows)
train <- subset(data_filtered, 1:nrows <= index_division)
test <- subset(data_filtered, 1:nrows > index_division)
actuals <- test$log_UMCSENT

models_list <- list(
  "ARIMA(0,1,3)" = c(0, 1, 3),
  "ARIMA(1,1,2)" = c(1, 1, 2),
  "ARIMA(2,1,1)" = c(2, 1, 1),
  "ARIMA(0,1,0)" = c(0, 1, 0)
)

comparison_results <- data.frame(
  Model = character(),
  LjungBox_p = numeric(),
  Shapiro_p = numeric(),
  RMSE = numeric(),
  MAE = numeric(),
  MAPE = numeric(),
  stringsAsFactors = FALSE
)

cat("Starting comparison loop\n")

for (name in names(models_list)) {
  
  order_params <- models_list[[name]]
  cat("\nProcessing:", name, "...\n")
  
  fit_train <- Arima(train$log_UMCSENT, order = order_params)
  # Residuals Diagnostics
  # Ljung-Box (autocorrelation)
  lb_test <- checkresiduals(fit_train, plot = FALSE) 
  lb_p_value <- lb_test$p.value
  # Shapiro-Wilk (Normality)
  sw_test <- shapiro.test(residuals(fit_train))
  sw_p_value <- sw_test$p.value
  
  # Rolling window
  n_train <- nrow(train)
  n_test <- nrow(test)
  preds <- numeric(n_test)
  
  for (i in 1:n_test) {
    current_data <- full_series[1:(n_train + i - 1)]
    fit_temp <- Arima(current_data, order = order_params)
    preds[i] <- forecast(fit_temp, h = 1)$mean
  }
  
  # Errors
  rmse_val <- sqrt(mean((preds - actuals)^2))
  mae_val <- mean(abs(preds - actuals))
  mape_val <- mean(abs((preds - actuals)/actuals)) * 100
  
  # Table
  comparison_results <- rbind(comparison_results, data.frame(
    Model = name,
    LjungBox_p = lb_p_value,   # > 0.05 (White Noise) - good
    Shapiro_p = sw_p_value,    # > 0.05 (Normale) - good
    RMSE = rmse_val,           
    MAE = mae_val,
    MAPE = mape_val
  ))
}

# RMSE comparison
comparison_results <- comparison_results[order(comparison_results$RMSE), ]

cat(" FINAL COMPARISON TABLE (Sorted by RMSE) \n")
print(comparison_results)

best_simple_model <- comparison_results$Model[1]
cat("\nThe best model is:", best_simple_model, "\n")

ggplot(comparison_results, aes(x = reorder(Model, RMSE), y = RMSE, fill = Model)) +
  geom_bar(stat = "identity", width = 0.6) +
  coord_flip() +
  labs(title = "Model Comparison by RMSE (Rolling Forecast)",
       x = "Model",
       y = "RMSE") +
  theme_minimal() +
  theme(legend.position = "none")
```

### ARIMA (0,1,3)

```{r}

## ARIMA(0,1,3)

library(forecast)
library(ggplot2)

nrows <- nrow(data_filtered)
index_division <- floor(0.80 * nrows)

train <- subset(data_filtered, 1:nrows <= index_division)
test <- subset(data_filtered, 1:nrows > index_division)

cat("Training set size:", nrow(train), "\n")
cat("Test set size:", nrow(test), "\n")

model_final <- Arima(train$log_UMCSENT, order = c(0, 1, 3))
print(summary(model_final))

checkresiduals(model_final)
# Shapiro Test
shapiro_test <- shapiro.test(residuals(model_final))
cat("Shapiro-Wilk test p-value:", shapiro_test$p.value, "\n")

library(ggplot2)
res_data <- data.frame(Residuals = as.numeric(residuals(model_final)))
# QQ-plot
ggplot(res_data, aes(sample = Residuals)) +
  stat_qq(color = "blue", alpha = 0.5, size = 2) +
  stat_qq_line(color = "red", size = 1) +           
  labs(title = "Normal Q-Q Plot of Residuals",
       subtitle = "Blue dots should follow the red line for normality",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()

# 4. ROLLING WINDOW FORECAST
cat("\nStarting Rolling Forecast\n")

full_series <- data_filtered$log_UMCSENT
n_train <- nrow(train)
n_test <- nrow(test)

rolling_preds <- numeric(n_test)

for (i in 1:n_test) {
  
  current_data <- full_series[1:(n_train + i - 1)]
  fit_temp <- Arima(current_data, order = c(0, 1, 3)) 
  rolling_preds[i] <- forecast(fit_temp, h = 1)$mean
  if(i %% 20 == 0) cat(".")
}
cat("\nRolling Forecast Completed.\n")
# ---------------------------------------------------------

# RMSE 
actuals <- test$log_UMCSENT
rmse_val <- sqrt(mean((rolling_preds - actuals)^2))
mae_val <- mean(abs(rolling_preds - actuals))
mape_val <- mean(abs((rolling_preds - actuals)/actuals)) * 100

cat("\n--- Rolling Forecast Accuracy ---\n")
cat("RMSE:", rmse_val, "\n")
cat("MAE :", mae_val, "\n")
cat("MAPE:", mape_val, "%\n")

# Plot
df_plot_train <- data.frame(
  Date = train$observation_date,
  Value = train$log_UMCSENT,
  Type = "Training Data"
)

df_plot_test <- data.frame(
  Date = test$observation_date,
  Value = test$log_UMCSENT,
  Type = "Actual Test Data"
)

df_plot_forecast <- data.frame(
  Date = test$observation_date,
  Value = rolling_preds, 
  Type = "Rolling Forecast"
)

df_total <- rbind(df_plot_train, df_plot_test, df_plot_forecast)

ggplot(df_total, aes(x = Date, y = Value, color = Type)) +
  geom_line(size = 0.8) +
  scale_color_manual(values = c("Training Data" = "gray60", 
                                "Actual Test Data" = "black", 
                                "Rolling Forecast" = "red")) +
  # Linea verticale per separare visivamente train e test
  geom_vline(xintercept = min(test$observation_date), linetype="dashed", color="blue", alpha=0.5) +
  labs(title = "ARIMA(2,1,3) Rolling Forecast vs Actual Data",
       subtitle = paste("One-step ahead prediction | RMSE:", round(rmse_val, 4)),
       y = "Log(Consumer Sentiment)",
       x = "Date") +
  theme_minimal() +
  theme(legend.position = "bottom")

```

### ARIMA (1,1,2)

```{r}


## 8. Training and Testing with ARIMA(1,1,2)

library(forecast)
library(ggplot2)

# 1. Split Dataset (Il codice che hai fornito)
nrows <- nrow(data_filtered)
index_division <- floor(0.80 * nrows)

train <- subset(data_filtered, 1:nrows <= index_division)
test <- subset(data_filtered, 1:nrows > index_division)

cat("Training set size:", nrow(train), "\n")
cat("Test set size:", nrow(test), "\n")

# 2. Costruzione del Modello ARIMA(0, 1, 3)
# p=2 (AR), d=1 (Integration/Difference), q=5 (MA)
model_final <- Arima(train$log_UMCSENT, order = c(1, 1, 2))

# Mostriamo il sommario del modello (Coefficienti, AIC, BIC)
print(summary(model_final))

# 3. Analisi dei Residui
# È fondamentale controllare che i residui siano "White Noise" (non correlati)
checkresiduals(model_final)

# Test di normalità sui residui
shapiro_test <- shapiro.test(residuals(model_final))
cat("Shapiro-Wilk test p-value:", shapiro_test$p.value, "\n")
# Se p-value > 0.05, i residui sono normali.

library(ggplot2)

# Estrai i residui dal modello
res_data <- data.frame(Residuals = as.numeric(residuals(model_final)))

# Crea il QQ-Plot
ggplot(res_data, aes(sample = Residuals)) +
  stat_qq(color = "blue", alpha = 0.5, size = 2) + # I punti dei residui
  stat_qq_line(color = "red", size = 1) +          # La linea della distribuzione normale teorica
  labs(title = "Normal Q-Q Plot of Residuals",
       subtitle = "Blue dots should follow the red line for normality",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()

# 4. ROLLING WINDOW FORECAST
# ---------------------------------------------------------
cat("\nStarting Rolling Forecast (this may take a few seconds)...\n")

full_series <- data_filtered$log_UMCSENT
n_train <- nrow(train)
n_test <- nrow(test)

rolling_preds <- numeric(n_test)

for (i in 1:n_test) {
  current_data <- full_series[1:(n_train + i - 1)]
  fit_temp <- Arima(current_data, order = c(1, 1, 2)) 
  
  # Previsione a 1 passo (il mese successivo)
  rolling_preds[i] <- forecast(fit_temp, h = 1)$mean
  
  # Barra di progresso semplice
  if(i %% 20 == 0) cat(".")
}
cat("\nRolling Forecast Completed.\n")
# ---------------------------------------------------------

# 5. Valutazione Accuratezza (Manuale per vettori)
# Calcoliamo RMSE tra le previsioni rolling e i dati reali del test set
actuals <- test$log_UMCSENT
rmse_val <- sqrt(mean((rolling_preds - actuals)^2))
mae_val <- mean(abs(rolling_preds - actuals))
mape_val <- mean(abs((rolling_preds - actuals)/actuals)) * 100

cat("\n--- Rolling Forecast Accuracy ---\n")
cat("RMSE:", rmse_val, "\n")
cat("MAE :", mae_val, "\n")
cat("MAPE:", mape_val, "%\n")

# 6. Visualizzazione Grafica: Training + Forecast vs Real Data
df_plot_train <- data.frame(
  Date = train$observation_date,
  Value = train$log_UMCSENT,
  Type = "Training Data"
)

df_plot_test <- data.frame(
  Date = test$observation_date,
  Value = test$log_UMCSENT,
  Type = "Actual Test Data"
)

df_plot_forecast <- data.frame(
  Date = test$observation_date,
  Value = rolling_preds, 
  Type = "Rolling Forecast"
)

# Uniamo tutto
df_total <- rbind(df_plot_train, df_plot_test, df_plot_forecast)

# Plot con ggplot2
ggplot(df_total, aes(x = Date, y = Value, color = Type)) +
  geom_line(size = 0.8) +
  scale_color_manual(values = c("Training Data" = "gray60", 
                                "Actual Test Data" = "black", 
                                "Rolling Forecast" = "red")) +
  # Linea verticale per separare visivamente train e test
  geom_vline(xintercept = min(test$observation_date), linetype="dashed", color="blue", alpha=0.5) +
  labs(title = "ARIMA(2,1,3) Rolling Forecast vs Actual Data",
       subtitle = paste("One-step ahead prediction | RMSE:", round(rmse_val, 4)),
       y = "Log(Consumer Sentiment)",
       x = "Date") +
  theme_minimal() +
  theme(legend.position = "bottom")

```

## 11. Residual Diagnostics

```{r}

## 7. Analisi dei Residui del Test Set (Forecast Errors)
# ---------------------------------------------------------
cat("\n--- Analisi Approfondita dei Residui di Previsione (Test Set) ---\n")

# 1. Calcolo dei residui (Reale - Previsto)
test_residuals <- test$log_UMCSENT - rolling_preds
df_test_res <- data.frame(Date = test$observation_date, Residuals = test_residuals)

# 2. Statistiche Descrittive
mean_res <- mean(test_residuals)
sd_res <- sd(test_residuals)

cat("Media dei residui (Bias):", mean_res, "(Ideale: vicino a 0)\n")
cat("Deviazione Standard:", sd_res, "\n")

# T-test per vedere se la media è statisticamente diversa da 0 (Bias check)
t_test <- t.test(test_residuals)
cat("T-test p-value (H0: mean=0):", t_test$p.value, "\n")
# Se p-value < 0.05, il modello ha un bias sistematico (sovrastima o sottostima costantemente).

# 3. Test di Ljung-Box sui residui del test
# Verifica se c'è ancora informazione (autocorrelazione) non catturata negli errori
lb_test <- Box.test(test_residuals, lag = 10, type = "Ljung-Box")
cat("Ljung-Box Test p-value:", lb_test$p.value, "\n")
# Se p-value > 0.05: Ottimo, gli errori sono "White Noise".
# Se p-value < 0.05: Male, c'è ancora autocorrelazione negli errori.

# 4. Test di Normalità (Shapiro-Wilk)
shapiro_test_res <- shapiro.test(test_residuals)
cat("Shapiro-Wilk Test p-value:", shapiro_test_res$p.value, "\n")

# ---------------------------------------------------------
# VISUALIZZAZIONE GRAFICA DEI RESIDUI DEL TEST
# ---------------------------------------------------------

# A. Time Plot dei Residui (per vedere se la varianza cambia nel tempo)
p1 <- ggplot(df_test_res, aes(x = Date, y = Residuals)) +
  geom_line(color = "darkblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Time Plot of Test Residuals",
       subtitle = "Residuals should fluctuate randomly around 0",
       y = "Residuals (Error)", x = "Date") +
  theme_minimal()
print(p1)

# B. ACF Plot dei Residui (Autocorrelazione)
# Usiamo ggAcf della libreria forecast
p2 <- ggAcf(test_residuals) +
  labs(title = "ACF of Test Residuals",
       subtitle = "Bars should stay within blue dashed lines (95% CI)") +
  theme_minimal()
print(p2)

# C. Istogramma con curva normale sovrapposta
p3 <- ggplot(df_test_res, aes(x = Residuals)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "skyblue", color = "white", alpha = 0.7) +
  stat_function(fun = dnorm, args = list(mean = mean_res, sd = sd_res), color = "red", size = 1) +
  labs(title = "Histogram of Test Residuals",
       subtitle = "Checking for Normal Distribution",
       x = "Residuals", y = "Density") +
  theme_minimal()
print(p3)

# D. QQ-Plot
p4 <- ggplot(df_test_res, aes(sample = Residuals)) +
  stat_qq(color = "blue", alpha = 0.5) +
  stat_qq_line(color = "red", size = 1) +
  labs(title = "QQ-Plot of Test Residuals", 
       subtitle = "Points should follow the red line",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()
print(p4)
```

```{r}


# AR ROOTS
library(forecast)
library(ggplot2)
library(gridExtra)

# --- 1. Fit AR Models ---
# We fit an AR(2) model to both series to see where the roots lie.
# We force d=0 because we want to see the properties of the raw data vs the manually differenced data.

# Model on Level Data (Log UMCSENT)
model_level_check <- Arima(data_filtered$log_UMCSENT, order = c(1, 0, 0))

# Model on Differenced Data
# We use na.omit because the first observation of the difference is NA
model_diff_check <- Arima(na.omit(data_filtered$diff_log_UMCSENT), order = c(1, 0, 0))

# --- 2. Plotting ---

# Plot A: Level Data
p1 <- autoplot(model_level_check) +
  labs(title = "Inverse AR Roots: Level Data",
       subtitle = "Log(UMCSENT) - Roots close to Unit Circle indicate Non-Stationarity")

# Plot B: Differenced Data
p2 <- autoplot(model_diff_check) +
  labs(title = "Inverse AR Roots: Differenced Data",
       subtitle = "First Diff - Roots inside Unit Circle indicate Stationarity")

# Display plots
print(p1)
print(p2)
```
